---
title: "Define new data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

## Defining live/dead trees

Status and condition codes provided do not necessarily define mortality, because a broken tree, depending on species may still be alive or dead. I'll define dead trees as:

1. snags or down snags (status -9 to -5, and 5 to 9)
2. trees broken below dbh and uprooted trees (status 30, 32)
3. any non-sprouting species broken above dbh (status 31)
4. any sprouting species broken below 3 meters
5. missing live trees


```{r define-live}

d_l <- d_l %>%
  mutate(
    live = case_when(
      status %in% c(5:9, -5:-9, -1)                                  ~ FALSE,
      status %in% c(30, 32)                                      ~ FALSE,
      (spp %in% c("PSMEM", "PISI") & status == 31)               ~ FALSE,
      (spp %in% c("SESE3", "ALRU") & status == 31 & ht < 3)      ~ FALSE,
      (spp %in% c("SESE3", "ALRU") & status == 31 & lag(ht) < 3) ~ FALSE,
      TRUE                                                       ~ TRUE
    )
  )

# Do any trees go from being dead to being alive?
nrow(filter(group_by(d_l, tree_id), any(!live & lead(live))))

```

## Define diameter, basal area, and height increment

Here I will add annual height, diameter and basal area increments to the dataset. Increment will be defined on the observation level, meaning that it must apply to either the first or second measurement. This is significant for modeling, as I will be predicting an increment based on one period's dbh observation. I'll add increment data to both the beginning and ending observations to facilitate modeling based on either observation. *_inc1 will be assigned to the first measurement (refers to subsequent measurement) and *_inc2 will be assigned to the second (refers to previous measurement).

```{r}
# Add diameter increment data and update post-harvest data
d_l <- d_l %>%
  group_by(tree_id) %>%
  mutate(
    ba = pi * (dbh / 2)^2,
    d_inc2 = (dbh - lag(dbh, order_by = year)) / 5,
    ht_inc2 = (ht - lag(ht, order_by = year)) / 5,
    ba_inc2 = (ba - lag(ba, order_by = year)) / 5,
    d_inc1 = (lead(dbh, order_by = year) - dbh) / 5,
    ht_inc1 = (lead(ht, order_by = year) - ht) / 5,
    ba_inc1 = (lead(ba, order_by = year) - ba) / 5
  ) %>%
  ungroup()
```


## Define UTM coordinates for each tree

```{r}

define_xy <- function(data) {
  data %>%
  left_join(plot_info[, c("plot", "utm_x", "utm_y")], by = "plot") %>%
  mutate(
    circular2xy(h_dist, azi, center.x = utm_x, center.y = utm_y)
  ) %>%
  select(-c(utm_x, utm_y))
}

d_l <- define_xy(d_l)

```

## Structural diversity SCI

After some experimentation, I determined that ingrowth does have an effect on SCI, especially for the more intense thinnings, supposedly because they let in more light and encourage more understory initiation. 

I also confirmed that the SCI function works the same regardless of whether coordinates are plot coordinates or UTM coordinates.

First, here is my code for defining SCI. The area of a triangle is equal to 1/2 the magnitude of the cross product of two vectors representing sides of a triangle. The magnitude of a vector is the square root of the the sum of its values squared. This is also known as the norm and is denoted with single or double vertical bars.

```{r}

cross_product <- function(a, b) {
  if(length(a)!=3 || length(b)!=3){
        stop("Cross product is only defined for 3D vectors.")
    }
  i1 <- c(2, 3, 1)
  i2 <- c(3, 1, 2)
  a[i1] * b[i2] - a[i2] * b[i1]
}

# area is the magnitude of two vectors
tri_area <- function(tri, points) {
  apply(tri, 1, function(point) {
    AB <- points[point[2], ] - points[point[1], ]
    AC <- points[point[3], ] - points[point[1], ]
    0.5 * sqrt(sum(cross_product(AB, AC)^2))
  })
}

sci_metric <- function(x, y, z) {
  points <- as.matrix(cbind(x, y, z))
  deln_obj <- geometry::delaunayn(points[, 1:2], output.options = "Fa")
  area3d <- sum(tri_area(deln_obj$tri, points))
  area2d <- sum(deln_obj$areas)
  area3d / area2d
}
```

Next, these functions calculate plot averages for SCI

```{r}
# calculate within plot coordinates, plot center = origin
# this does not work with missing values currently.

# confidence interval for average treatment estimate
ci <- function(x) {
  LB <- mean(x) - 2 * sd(x) / sqrt(length(x) - 1)
  UB <- mean(x) + 2 * sd(x) / sqrt(length(x) - 1)
  c(LB, UB)
}

calc_sci <- function(data, metric) {
  data %>%
    group_by(treatment, year, plot) %>%
    mutate(
      "sci_{{metric}}" := sci_metric(x, y, {{metric}}),
    )
}

summary_sci <- function(data, level = treatment) {
  plot_level <- data %>%
  calc_sci(dbh) %>%
  summarise(across(starts_with("sci"), first))
  if(level == plot) {
    plot_level
  } else {
    plot_level %>% summarise(
      across(
        starts_with("sci"),
        list(avg = mean, ci_lb = ~ ci(.x)[1], ci_ub = ~ ci(.x)[2])
      )
    )
  }
}
```

## Standardize crown ratio

```{r}
d_l <- d_l %>%
  mutate(
    cr = case_when(
      cr %in% c(-999, -99) ~ NA_real_,
      cr == 0 & is.na(ht)  ~ NA_real_,
      cr %in% 1:5          ~ 5,
      cr %in% 21:25        ~ 25,
      cr %in% 31:35        ~ 35,
      cr %in% 41:50        ~ 50,
      cr %in% 55:80        ~ 60,
      TRUE                 ~ cr
    )
  )
```

## Calculate plot SDI

For "live" trees only

```{r}

d_l <- d_l %>%
  group_by(plot) %>%
  mutate(sdi_plot = sum(12.5 * (dbh * live / 25.4)^1.605, na.rm = TRUE)) %>%
  ungroup()

```

## Update post-harvest data

Make sure that this is the last step

```{r}
p_h <- filter(d_l, year > "init")
```
