---
title: "Modeling height growth and predicting missing heights"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

```{r}
library(lme4)
library(emmeans)
library(nlme)
library(MuMIn)
library(emmeans)
library(ggpubr)
palette("Tableau 10")
```

# Introduction

I need to predict missing heights in 2018 in order to complete other steps of the analysis and provide better summary data. To accomplish this, I will model height increment as a function of the continuous predictor, dbh, as well as a combination of nested groupings: year, treatment, plot, and species. Additionally, height increment response is of interest in it's own right. While height growth in trees is less responsive to conditions then diameter, it would be important to see whether there is a difference detected between treatments.

# Defining trees of interest

First, I'll define the dataset of interest as only years 2013 and 2018 observations of healthy, non-leaning, unbroken SESE or PSME trees. I'll also define three other alternative treatment groupings for consideration: thinned/unthinned, H/L/C (thinning type), and 40/80/C (thinning intensity).

I will predict height increment based on the beginning of period dbh. Predicting based on the end of period dbh resulted in a lower AIC, but I am defaulting to typical practice. 

I'm also using height increment from both 2008 - 2013 and 2013 to 2018.

```{r}

# test_d2 <- d_l %>%
#   group_by(tree_id) %>%
#   filter(
#     spp %in% c("SESE3", "PSMEM"),
#     year %in% c("13", "18"),
#     live,
#     !get_cond(2, 3, 5),
#     !is.na(ht_inc2) & !is.na(dbh)
#   ) %>%
#   mutate(
#     treatment2 = str_extract(treatment, "C|H|L"),
#     treatment3 = str_extract(treatment, "C|40|80"),
#     treatment4 = if_else(str_detect(treatment, "C"), "unthinned", "thinned"),
#     year = factor(year, levels = c("13", "18"), ordered = FALSE)
#   ) %>%
#   select(starts_with("treatment"), spp, year, tree_id, dbh, ht_inc2, plot, x, y)

test_d <- d_l %>%
  group_by(tree_id) %>%
  filter(
    spp %in% c("SESE3", "PSMEM"),
    year %in% c("08", "13"),
    lead(live),
    !lead(get_cond(2, 3, 5)),
    !is.na(ht_inc1) & !is.na(dbh),
  ) %>%
  ungroup() %>%
  mutate(
    treatment2 = str_extract(treatment, "C|H|L"),
    treatment3 = str_extract(treatment, "C|40|80"),
    treatment4 = if_else(str_detect(treatment, "C"), "unthinned", "thinned"),
    year = factor(year, levels = c("08", "13"), ordered = FALSE)
  ) %>%
  select(starts_with("treatment"), spp, year, tree_id, dbh, ht_inc1, plot, x, y)

# test_d12 <- d_l %>%
#   group_by(tree_id) %>%
#   filter(
#     spp %in% c("SESE3", "PSMEM"),
#     year %in% c("13"),
#     lead(live),
#     !lead(get_cond(2, 3, 5)),
#     !is.na(ht_inc1) & !is.na(dbh),
#   ) %>%
#   ungroup() %>%
#   mutate(
#     treatment2 = str_extract(treatment, "C|H|L"),
#     treatment3 = str_extract(treatment, "C|40|80"),
#     treatment4 = if_else(str_detect(treatment, "C"), "unthinned", "thinned"),
#     year = factor(year, levels = c("13"), ordered = FALSE)
#   ) %>%
#   select(starts_with("treatment"), spp, year, tree_id, dbh, ht_inc1, plot, x, y)

```

# Assessment of linear relationship between dbh and height increment

We are pretty sure that height growth is correlated with diameter, in that larger trees are capable of more height growth than smaller ones. We also think that this relationship is probably not linear. I will take a look at log and square root transformations of dbh for linear prediction of height growth.

```{r, fig.height=7, fig.width=10}
m0 <- lm(ht_inc1 ~ dbh * treatment * spp, data = test_d)
m1 <- update(m0, ~ sqrt(dbh) * treatment * spp)
m2 <- update(m0, ~ log(dbh) * treatment * spp)


test_d %>%
  ggplot(aes(x = dbh, y = ht_inc1, color = spp)) +
    geom_point(aes(color = spp), alpha = .5) +
    facet_wrap(vars(treatment)) +
    geom_line(aes(y = predict(m0), linetype = "linear"), size = 1) +
    geom_line(aes(y = predict(m1), linetype = "sqrt"), size = 1.25) + 
    geom_line(aes(y = predict(m2), linetype = "log"), size = 1.25) + 
    scale_color_manual(
      values = palette(),
      name = "",
      breaks = c("PSMEM", "SESE3", "linear", "sqrt", "log"),
    )

rm(m0, m1, m2)
```

It doesn't look like it matters a whole lot with this noisy data. I'm going to use log transformation.

# Global Model

The next step is going to specify our likely global model. In general, we want to answer questions about how a given treatment affected growth for each species. There may have also been different responses between years. This will be a mixed model and I want to control for the random effect of plot and tree_id. The global model will include `log(dbh)` as well as the categories: `treatment`, `species`, `year`, and all their interactions. Resulting in the fixed effects:

`ht_inc1 ~ log(dbh) * treatment * spp * year`

# Optimal random effects specification

We will determine the optimal random effects structure using the global fixed effects structure. I will select the best of the following random effects structures:

    (1 | plot)
    (1 | plot) + (1 | tree_id)
    (log(dbh) | plot) + (1 | tree_id)

I'll do this using AICc and Restricted Maximum Likelihood Estimation (REML), which I believe is the preferred method when comparing random effects structure, whereas ML is necessary when comparing different fixed effects.

```{r ranef-spec}
f0 <- formula(ht_inc1 ~ log(dbh) * treatment * spp * year)
m0 <- gls(f0, data = test_d, method = "REML")
m1 <- lmer(update(f0, ~ . + (1 | plot)), REML = TRUE, data = test_d)
m2 <- update(m1, ~ . + (1 | tree_id))
m3 <- update(m2, ~ . + (0 + log(dbh) | plot))

AICc(m0, m1, m2, m3)
```

The optimal random effects structure, includes the random intercept only for `plot`. I will assume this random effects structure in all further height models.

# Choosing a sub-model for fixed effects

The next step is to define a list of potential sub-models for fixed effects to determine the optimal structure. 

```{r}

fl <- list(
  ht_inc1 ~ log(dbh) + (1 | plot),
  ht_inc1 ~ log(dbh) + spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment + spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp + treatment + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment * spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp * treatment * year + (1 | plot),
  ht_inc1 ~ log(dbh) + spp * year + treatment + (1 | plot),
  ht_inc1 ~ log(dbh) + spp + treatment * year + (1 | plot),
  ht_inc1 ~ log(dbh) * treatment * year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment + year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment2 + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment2 + spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp + treatment2 + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment2 * spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp * treatment2 * year + (1 | plot),
  ht_inc1 ~ log(dbh) + spp * year + treatment2 + (1 | plot),
  ht_inc1 ~ log(dbh) + spp + treatment2 * year + (1 | plot),
  ht_inc1 ~ log(dbh) * treatment2 * year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment2 + year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment3 + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment3 + spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp + treatment3 + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment3 * spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp * treatment3 * year + (1 | plot),
  ht_inc1 ~ log(dbh) + spp * year + treatment3 + (1 | plot),
  ht_inc1 ~ log(dbh) + spp + treatment3 * year + (1 | plot),
  ht_inc1 ~ log(dbh) * treatment3 * year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment3 + year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment4 + (1 | plot),
  ht_inc1 ~ log(dbh) * treatment4 + spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp + treatment4 + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment4 * spp + (1 | plot),
  ht_inc1 ~ log(dbh) * spp * treatment4 * year + (1 | plot),
  ht_inc1 ~ log(dbh) + spp * year + treatment4 + (1 | plot),
  ht_inc1 ~ log(dbh) + spp + treatment4 * year + (1 | plot),
  ht_inc1 ~ log(dbh) * treatment4 * year + spp + (1 | plot),
  ht_inc1 ~ log(dbh) + treatment4 + year + spp + (1 | plot)
)

```

I use AICc to assess the fit of each of the models with coefficients estimated with ML in order to compare among various fixed effects.

```{r}

get_aic(fl, data = test_d) %>%
  kbl(caption = "AICS for ht increment models") %>%
  kable_styling(full_width = FALSE)

lmod <- function(fl) {
  lmer(fl, data = test_d, REML = TRUE)
}

summary(lm(ht_inc1 ~ log(dbh) + treatment4 * spp, data = test_d))
summary(lmod(fl[[32]]))
```

# Best model (scrapping above)

While these models are interesting, there is reason to believe that our height measurements are not precise enough to capture the small height differences between treatments (around 3 feet). For this reason, I will default to a simple additive model that includes spp and treatment.

```{r}

ht_inc_mod1 <- lmod(fl[[32]])

summary(ht_inc_mod1)


```

Here is the summary table of the model used for predicting height increments and thus heights

```{r}
library(broom.mixed)
library(huxtable)

tidy(ht_inc_mod1)

huxreg("Height increment" = ht_inc_mod1, error_pos = "right", statistics = c("N" = "nobs"), stars = NULL)

sjPlot::tab_model(ht_inc_mod1, dv.labels = "Height increment")

modelsummary::modelsummary(ht_inc_mod1)
```

## Best model estimated means

switch to model 32 

```{r}
# changed model here treatment4
emmip(
  lmod(fl[[32]]),
  spp ~ dbh | treatment4,
  CIs = TRUE,
  plotit = FALSE,
  at = list(dbh = seq(10, 60, 2))
) %>% 
  ggplot(aes(dbh, yvar, color = spp)) +
    geom_line() +
    facet_wrap(~ treatment4 ) +
    geom_ribbon(aes(ymin = LCL, ymax = UCL, color = NULL, fill = spp), alpha = .2) +
    scale_color_manual(values = palette(), aesthetics = c("color", "fill")) +
    labs(y = "Predicted ht increment")

emmip(
  lmod(fl[[5]]),
  spp ~ dbh | treatment,
  CIs = TRUE,
  plotit = FALSE,
  at = list(dbh = seq(10, 60, 2))
) %>% 
  ggplot(aes(dbh, yvar, color = spp)) +
    geom_line() +
    facet_wrap(~ treatment ) +
    geom_ribbon(aes(ymin = LCL, ymax = UCL, color = NULL, fill = spp), alpha = .2) +
    scale_color_manual(values = palette(), aesthetics = c("color", "fill")) +
    labs(y = "Predicted ht increment")

emmeans(ht_inc_mod1, pairwise ~ treatment4, by = c("spp"))
```

## Best model validation

```{r}
# augment data with fitted, residual cooks distance and leverage

augment1 <- function(dat, mod) {
  dat %>% mutate(
    fitted = fitted(mod),
    resid = resid(mod, type = "pearson", scaled = TRUE),
    cooks = cooks.distance(mod),
    lev = hatvalues(mod)
  )
}

ht_inc_d <- augment1(test_d, ht_inc_mod1)

```

### Residual vs fitted

I also plot residual vs the predictor

```{r}
par(mfrow = c(1, 2))
plot(
  resid ~ fitted,
  data = ht_inc_d,
  pch = 16,
  xlab = "fitted values",
  ylab = "Scaled residuals",
  main = "Residual vs fitted",
  col = 2
)
abline(0,0)
plot(
  resid ~ log(dbh),
  data = ht_inc_d,
  xlab = "log(dbh)",
  ylab = "Scaled residuals",
  main = "Residual vs log(dbh)",
  col = 2,
  pch = 16
)
abline(0,0)

with(
  model.frame(ht_inc_mod1),
  boxplot(
    resid(ht_inc_mod1, type = "pearson") ~ spp,
    xlab = "by spp",
    ylab = "Residuals",
    main = "Distribution of residuals by spp"
  )
)

```

### Homogeneity of random group residuals

I can check that random group residuals are homogenous

```{r}
plot(
  ht_inc_mod1,
  resid(., scaled=TRUE) ~ fitted(.)| plot,
  abline = 0,
  pch = 16,
  xlab = "Fitted values",
  ylab = "Standardised residuals"
)
```


### Normality of residuals

Checking for normality of residuals. The tails are perhaps a bit fat, I'm not sure if this requires attention or not.

```{r, fig.height=7, fig.width=10}
qqnorm(resid(ht_inc_mod1, type = "pearson"), pch=16, col = 1,  main = "QQplot for pearosn residuals")
qqline(resid(ht_inc_mod1, type = "pearson"))
```

### Leverage and Cooks distance

Cooks outliers, defined as > 3 x mean(cooks distance) are colored in red. None of these "outliers" seem like they would disproportionately affect regression. 

```{r, fig.width=10, fig.height=8}
# Plot leverage against standardised residuals
plot(
  resid ~ lev,
  data = ht_inc_d,
  las = 1,
  ylab = "Standardised residuals",
  xlab = "Leverage",
  col = palette.colors(palette = "tableau10", alpha = .5)[1],
  main = "Leverage vs residuals",
  pch = 16
)
points(resid ~ lev, data = filter(ht_inc_d, cooks > 3 * mean(cooks)), pch = 16, col =3)
```

### Random effects distribution

The distribution of random effects (plots) should be roughly normal. Here our distribution is skewed, [from what I've read], this should not be a problem for estimates or their std's.

[from what I've read]: (https://stats.stackexchange.com/a/366500)

```{r}
hist(as.vector(unlist(ranef(ht_inc_mod1)$plot)), col = 1)
```

# Update data with predicted heights 

I will start by predicting heights for Douglas-fir and Redwood only. I can only predict heights for trees in year 13 and 18, and only if they have a height in the previous period. I can predict heights for trees whose crown status did not change from one measurement to the next. I will be assuming that a broken tree resumes height growth at a "normal" rate.

I'll predict heights for all trees and then filter this to include only observations of live trees that didn't break, or get a dead top (or start leaning) from one observation to the next.

```{r}

new_heights <- d_l %>%
  mutate(treatment4 = if_else(str_detect(treatment, "C"), "unthinned", "thinned")) %>%
  filter(spp %in% c("SESE3", "PSMEM")) %>%
  mutate(pred_ht_inc = predict(ht_inc_mod1, newdata = .)) %>%
  group_by(tree_id) %>%
  mutate(
    pred_ht = lag(ht) + lag(pred_ht_inc) * 5,
    cond_chk = get_cond(2, 3, 5, 30, 31, 32, str = TRUE),
    ht_p = case_when(
      cond_chk == lag(cond_chk) & live & is.na(ht) ~ pred_ht,
      TRUE ~ ht
    )
  ) %>%
  select(tree_id, year, ht_p) %>%
  ungroup()

```

Now I'll update the data with height predictions for selected trees and fill in available heights for other species that were not predicted.

```{r}

if ("ht_p" %in% names(d_l)) d_l["ht_p"] <- NULL

d_l <- left_join(d_l, new_heights, by = c("tree_id", "year")) %>%
  mutate(ht_p = if_else(spp %in% c("SESE3", "PSMEM"), ht_p, ht))


```
